Validation of an Actuator Line Model Coupled to a Dynamic Stall Model for Pitching Motions Characteristic to Vertical Axis Turbines

### 2016-06-13 paper.tex
'rel' in subscript should not be italics
Last modifications
Unsteady effects chapter correction,
Force distribution subchapter added
Others minimal changes VAWT and HAWT

\begin{abstract}
    Vertical axis turbines (VATs) can be used to extract renewable energy from
    wind or water flows. A simpler design, low cost of maintenance, and the
    ability to accept flow from all directions perpendicular to the rotor axis
    are some of the most important advantages over conventional horizontal axis
    turbines (HATs). However, VATs encounter complex and unsteady fluid
    dynamics, which present significant modeling challenges. One of the most
    relevant phenomena is dynamic stall, which is caused by the unsteady
    variation of angle of attack throughout the blade rotation, and is the focus
    of the present study. Dynamic stall is usually used as a passive control for
    VAT operating conditions, hence the importance of predicting its effects. In
    this study, a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
    line model and dynamic stall model are used to compute the blade loading and
    body force. Force coefficients obtained from the model are validated with
    experimental data of pitching airfoil in similar operating conditions as an
    H-rotor type VAT. Numerical results show reasonable agreement with
    experimental data for pitching motion.
\end{abstract}

@@

Operating conditions of Vertical Axis Wind Turbines VAWTs are characterized as
complex unsteady flows which give a considerable challenge, both to describe
using measurements and to represent through simulation
tools~\cite{huyer1996unsteady}. Moreover, VAWT blades are inherently exposed to
cyclic variation in the angle of attack, which gives cyclic blade forces and can
give material fatigue damage. Accurate modeling of the varying forces is
therefore very important for the design of the VAWT.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{rel} \right|^2  \label{lift}
\end{align} %\begin{align}
\begin{align}
& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{rel} \right|^2  \label{drag}
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are
function of the Reynolds number and the angle of attack. It should be emphasized
that the direction of the lift and drag are respectively perpendicular and
parallel to the relative velocity $\vec{V}_{\mathrm{rel}}$. $c$ represents the
chord length.

@@

@@

\subsection{Unsteady effects correction}
A test for the simulation with the DSM turned off was carried out, it means there is no compensation for the unsteady effects: the effective angle of attack is the angle of attack provided by the ALM. Figure \ref{AMP174corrections} depicts that simulated results are closer to experimental data with the corrections. Both $C_N$ and $C_T$ peak values are not overestimated. The $C_N$ loop, shows that the model without the corrections was not able to capture reattachment flow effects.

@@

It has been demonstrated that the presented coupled model ALM with DSM, shows
good agreement with experimental data, considering the complexity of the flows
studied. Moreover, the model has been tested for a wide range of pithing
amplitudes, therefore, with different stall conditions (none, shallow, and deep
stall), results agreed well compared with experiments. The model is stable and
accurate for the tested pithing blade motions, which makes it suitable for
application in VAT simulations.

>>>>

\begin{abstract}
    Vertical axis wind turbines (VAWT) can be used to extract renewable energy from
    wind flows. A simpler design, low cost of maintenance, and the
    ability to accept flow from all directions perpendicular to the rotor axis
    are some of the most important advantages over conventional horizontal axis wind
    turbines (HAWT). However, VAWT encounter complex and unsteady fluid
    dynamics, which present significant modeling challenges. One of the most
    relevant phenomena is dynamic stall, which is caused by the unsteady
    variation of angle of attack throughout the blade rotation, and is the focus
    of the present study. Dynamic stall is usually used as a passive control for
    VAWT operating conditions, hence the importance of predicting its effects. In
    this study, a coupled model is implemented with the open-source CFD
    toolbox OpenFOAM for solving the Navier--Stokes equations, where an actuator
    line model and dynamic stall model are used to compute the blade loading and
    body force. Force coefficients obtained from the model are validated with
    experimental data of pitching airfoil in similar operating conditions as an
    H-rotor type VAWT. Numerical results show reasonable agreement with
    experimental data for pitching motion.
\end{abstract}

@@

Operating conditions of Vertical Axis Wind Turbines (VAWT) are characterized as
complex unsteady flows which give a considerable challenge, both to describe
using measurements and to represent through simulation
tools~\cite{huyer1996unsteady}. Moreover, VAWT blades are inherently exposed to
cyclic variation in the angle of attack, which gives cyclic blade forces and can
give material fatigue damage. Accurate modeling of the varying forces is
therefore very important for the design of the VAWT.

@@

Once the angle of attack, relative velocity are obtained, the blade element lift and drag forces per spanwise length unit are calculated as
\begin{align}
& f_L = \frac{1}{2} \ \rho \ c \ C_L \left| V_{\mathrm{rel}} \right|^2  \label{lift}
\end{align} %\begin{align}
\begin{align}
& f_D = \frac{1}{2} \ \rho \ c \ C_D \left| V_{\mathrm{rel}} \right|^2  \label{drag}
\end{align} %\begin{align}
where $C_L$ and $C_D$ are the lift and drag coefficients, respectively. Both are
function of the Reynolds number and the angle of attack. It should be emphasized
that the direction of the lift and drag are respectively perpendicular and
parallel to the relative velocity $\vec{V}_{\mathrm{rel}}$. $c$ represents the
chord length.

@@

\section*{Calculated force distribution}
The applied forces in the ALM need to be distributed smoothly on several mesh points in order to avoid instability due to high gradients. Source term forces should be projected around the element location using a three-dimensional Gaussian kernel. This gives the smoothing function $\eta$, which is multiplied by the computed local force on the actuator line element and then imparted on a cell with a distance $| \vec r |$ from the actuator line element quarter chord location:

\begin{align*}
\eta = \frac{1}{\epsilon ^3 \pi ^{3/2} } \exp \left[ - \left( \frac{| \vec r |}{\epsilon} \right) ^2 \right]
\end{align*}

The smoothing width parameter $\epsilon$ is chosen by the maximum value from three contributions: one relative to the 25\% of the chord length, one to the mesh size, and one to the momentum thickness due to drag force, and it is expressed as:

\begin{align*}
\epsilon = \max \left[ \frac{c}{4},4 \sqrt[\leftroot{-2}\uproot{2}3]{V_{\mathrm{cell}}},  \frac{c C_d}{2}\right]
\end{align*}

Where ${V_{\mathrm{cell}}}$ is the cell volume. More details about force projection are in \cite{Bachant2016-VAT-ALM}.

@@

\subsection{Unsteady effects correction}

According to \cite{dyachuk2015simulating}, where the dynamic stall model is combined with a vortex model, it is beneficial to omit the attached flow correction equation \ref{alphaeq} and instead use $\alpha_E$ = $\alpha_n$. Hence, it is of interest to investigate if this correction should be included or not when the ALM is used instead of the vortex model. Figure \ref{AMP174corrections} depicts that simulated results with the attached flow corrections have small improvements in the accuracy and are closer to experimental data. Results without corrections also have good agreement. Both $C_N$ and $C_T$ peak values are not overestimated.  This same study was carried out for a bigger angle of attack range and differences are less significant.        This has also been tested for larger angles of attack and the differences between using equation \ref{alphaeq} or not are less significant here as the separated flow modeling dominates the force calculation.

@@

It has been demonstrated that the presented coupled model ALM with DSM, shows
good agreement with experimental data, considering the complexity of the flows
studied. Moreover, the model has been tested for a wide range of pithing
amplitudes, therefore, with different stall conditions (none, shallow, and deep
stall), results agreed well compared with experiments. The model is stable and
accurate for the tested pithing blade motions, which makes it suitable for
application in VAWT simulations.
